---
title: "The Good, the Bad and the Ugly"
author: "Jack Carter"
date: "18/04/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(ggthemes)
library(knitr)
library(here)

# gets the z score for each term. 
get_z_score <- function(data) {
  mean <- mean(data)
  sigma <- sd(data)
  z_scores <- list()
  for(i in 1:length(data)) {
    z_scores[i] <- (data[i] - mean) / sigma
  }
  return(unlist(z_scores))
}

# gets the z score for a list of terms. 
get_z_scores <- function(data) {
  groups <- data %>%
    group_split(term)
  z_scores <- list()
  for(i in 1:length(groups)) {
    z_scores[[i]] <- list(get_z_score(groups[[i]]$hits))
  }
  return(unlist(z_scores))
}

national_belonging <- c("patriotism",
                        "national identity")
partisanship <- c("partisanship",
                  "political polarization",
                  "political divide",
                  "political differences")
discrimination <- c("anti-semitism",
                    "sexism", 
                    "racism", 
                    "islamophobia",
                    "transphobia")

# read in data. 
data <- read_csv(here("data.csv"))

# sort data alphabetically by term. 
index <- order(data$term)
sorted_df <- data[index,]

# change term to upper case. 
sorted_df$term <- str_to_title(sorted_df$term)

# adds a z-score variable. 
sorted_df$z_score <- get_z_scores(sorted_df)

# my personal plot theme for data visualizations. 
my_theme <- theme_economist_white(gray_bg = FALSE) +
  theme(plot.title = element_text(hjust = 0.5,
                                  vjust = 15,
                                  size = 10,
                                  color = "#474747"),
  plot.margin = unit(c(1.5, 1, 1.5, 1), "cm"),
  axis.text = element_text(size = 9,
                           color = "gray30"),
  axis.text.x=element_text(vjust = -2.5),
  axis.title.x = element_text(size = 9,
                              color = "gray30",
                              vjust = -10),
  axis.title.y = element_text(size = 9,
                              color = "gray30",
                              vjust = 10),
  legend.direction = "vertical", 
  legend.position = "right",
  legend.title = element_blank(),
  legend.text = element_text(size = 11,
                             color = "gray20"),
  legend.margin=margin(1, -15, 1, 0),
  legend.spacing.x = unit(0.25, "cm"),
  legend.key.size = unit(1, "cm"), 
  legend.key.height = unit(0.75, "cm"),
  strip.text = element_text(hjust = 0.5,
                            vjust = 1,
                            size = 10,
                            color = "#474747"),
  panel.spacing = unit(2, "lines"))

# creates a plot with smoothed loess regression lines. 
make_plot <- function(category, title) {
  plot <- sorted_df %>%
    filter(term %in% str_to_title(category)) %>%
    ggplot(aes(x=year, 
               y=z_score, 
               col=term)) +
    geom_smooth(se=F, 
                span = 0.5, 
                size = 0.5) +
    geom_vline(xintercept = 2016,
               size=0.25,
               col="#696969") +
    geom_text(aes(x=2016, 
                  label="2016 win -", 
                  y=max(z_score)+0.1,
                  hjust=1.05),
              size=2.75,
              col="#696969") +
    geom_vline(xintercept = 2020,
               size=0.25,
               col="#696969") +
    geom_text(aes(x=2020, 
                  label="2020 loss -",
                  y=max(z_score)+0.1,
                  hjust=1.05),
              size=2.75,
              col="#696969") +
    ggtitle(title) +
    ylab("Articles (z-scores)") +
    xlab("") + 
    my_theme
  return(plot)
}

```

## **Summary**
This project shows how the conversation in the New York Times (NYT) changed following the 2016 election of Donald Trump. A strong sense of national belonging is undoubtedly good for any country, yet Trump's rise also appears to have been accompanied by growing political partisanship and increased social discrimination.

&nbsp;

## 1) The Good

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# national_belonging  
make_plot(national_belonging, "National Belonging")

```

## 2) The Bad

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# partisanship   
make_plot(partisanship, "Partisanship")

```

## 3) The Ugly

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# discrimination 
make_plot(discrimination, "Discrimination")

```

&nbsp;

## **Method**

### **1) Data Collection:**

The data were collected from the New York Times API. They include the number of articles for each search term in the 12 years between 2011 and 2022.

**Total articles (thousands)**

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# creates a summary table of the total hits for each term. 
summary_table <- data %>%
  mutate(term=str_to_title(term),
         hits=round((hits/10^3), 2)) %>%
  group_by(term) %>%
  summarise(total=sum(hits)) %>%
  spread(term, total)
kable(summary_table)

```

&nbsp;

### **2) Z-scores:**

The number of articles is converted to each term's z-score. This allows us to view the term's relative distribution over time. It is calculated as 1) the number of articles less the term's mean, 2) divided by the term's standard deviation.

—EXAMPLE CODE SNIPET—

```{r, echo = TRUE, message = FALSE, warning = FALSE, dpi=600}

# gets the z-score for each term. 
get_z_score <- function(data) {
  mean <- mean(data)
  sigma <- sd(data)
  z_scores <- list()
  for(i in 1:length(data)) {
    z_scores[i] <- (data[i] - mean) / sigma
  }
  return(unlist(z_scores))
}

# gets the z-score for a list of terms. 
get_z_scores <- function(data) {
  groups <- data %>%
    group_split(term)
  z_scores <- list()
  for(i in 1:length(groups)) {
    z_scores[[i]] <- list(get_z_score(groups[[i]]$hits))
  }
  return(unlist(z_scores))
}

```

&nbsp;

### **3) Loess Transformation:**

The data for each term is plotted with the use of a loess regression line (geom_smooth in the code below). This transforms the data into a smooth curve for a better visualization of underlying trends. 

—EXAMPLE CODE SNIPET—

```{r, echo = TRUE, message = FALSE, warning = FALSE, dpi=600}

# creates a plot with smoothed loess regression lines. 
make_plot <- function(category, title) {
  plot <- sorted_df %>%
    filter(term %in% str_to_title(category)) %>%
    ggplot(aes(x=year, 
               y=z_score, 
               col=term)) +
    geom_smooth(se=F, 
                span = 0.5, 
                size = 0.5) +
    geom_vline(xintercept = 2016,
               size=0.25,
               col="#696969") +
    geom_text(aes(x=2016, 
                  label="2016 win -", 
                  y=max(z_score)+0.1,
                  hjust=1.05),
              size=2.75,
              col="#696969") +
    geom_vline(xintercept = 2020,
               size=0.25,
               col="#696969") +
    geom_text(aes(x=2020, 
                  label="2020 loss -",
                  y=max(z_score)+0.1,
                  hjust=1.05),
              size=2.75,
              col="#696969") +
    ggtitle(title) +
    ylab("Articles (z-scores)") +
    xlab("") + 
    my_theme
  return(plot)
}

```

&nbsp;

## **Sources**

- New York Times (2021) https://developer.nytimes.com/apis

- Statology (2021) https://www.statology.org/interpret-z-scores/

&nbsp;
&nbsp;
&nbsp;

