---
title: "The Good, the Bad and the Ugly"
author: "Jack Carter"
date: "10/17/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(stringr)
library(ggthemes)
library(here)

# gets the z score for a specific term. 
get_z_score <- function(data) {
  mean <- mean(
    data
    )
  sigma <- sd(
    data
  )
  z_scores <- list()
  for(i in 1:length(data)) {
    z_scores[i] <- (data[i] - mean) / sigma
  }
  return(unlist(
    z_scores
    ))
}

# gets the z score for a list of terms. 
get_z_scores <- function(data) {
  groups <- data %>%
    group_split(term)
  
  z_scores <- list()
  for(i in 1:length(groups)) {
    z_scores[i] <- list(
      get_z_score(groups[[i]]$hits)
      )
  }
  return(unlist(
    z_scores
    ))
}

trump <- "Donald Trump"

national_belonging <- c("patriotism",
                        "nationalism",
                        "national identity")

disinformation <- c("conspiracy",
                    "disinformation",
                    "falsehoods")

media_distrust <- c("fake news", 
                    "enemy of the people",
                    "media manipulation")

partisanship <- c("partisan politics",
                  "political polarization",
                  "political divide")

discrimination <- c("discrimination",
                    "bigotry",
                    "hate speech")

racism <- c("racism",
            "white supremacy",
            "xenophobia")

sexism <- c("sexism",
            "misogyny",
            "transphobia")

sexual_orientation <- c("homophobia",
                        "transphobia")

far_right <- c("the far-right",
               "white nationalists",
               "white supremacists")

# read in data. 
data <- read_csv("data.csv")

# sort data alphabetically by term. 
index <- order(data$term)
sorted_data_frame <- data[index,]

# change term to upper case. 
sorted_data_frame$term <- str_to_title(sorted_data_frame$term)

# my personal plot theme for data visualizations. 
my_theme <- theme_economist_white(gray_bg = FALSE) +
  theme(plot.title = element_text(hjust = 0.5, 
                                  vjust = 13, 
                                  size = 8, 
                                  color = "#474747"),
        plot.margin = unit(c(1.5, 1, 1.5, 1), "cm"),
        axis.text = element_text(size = 9, 
                                 color = "gray30"),
        axis.text.x=element_text(vjust = -2.5),
        axis.title.x = element_text(size = 9, 
                                    color = "gray30", 
                                    vjust = -10),
        axis.title.y = element_text(size = 9, 
                                    color = "gray30", 
                                    vjust = 10),
        legend.direction = "vertical", 
        legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 10, 
                                   color = "gray20"),
        legend.margin=margin(1, -15, 1, 0),
        legend.spacing.x = unit(0.25, "cm"),
        legend.key.size = unit(1, "cm"), 
        legend.key.height = unit(0.75, "cm"),
        strip.text = element_text(hjust = 0.5, 
                                  vjust = 1, 
                                  size = 10, 
                                  color = "#474747"),
        panel.spacing = unit(2, "lines"))

```

## **Summary**
This project shows how the conversation in the New York Times (NYT) turned for the worse following the 2016 election of Donald Trump. A strong sense of national belonging is undoubtedly good for any country, yet Trump's rise was also accompanied by a growing distrust in the media and increased social divisiveness. While this data may indicate a decline in such rhetoric following Joe Biden's 2020 victory, the underlying views of many Americans who do not read or contribute to liberal newspapers like the NYT have likely not changed. 

&nbsp;

## **Method**

**1) Data Collection:**

The data used in this study are collected from the NYT API. They include metadata indicating the number of articles that contain certain terms for each year between 2011 and 2021. 

**2) Z Scores:**

The z scores are then calculated as the raw value less the term's mean, divided by the term's standard deviation. This allows us to compare the relative trajectory of terms regardless of their overall count during Trump's presidency. 

&nbsp;

## The Rise of Trump

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# trump's rise.  
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
    ) %>%
  filter(
    term %in% str_to_title(trump)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
    ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
    ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
    ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Trump's Rise"
    ) +
  ylab(
    "Z Score"
    ) +
  xlab(
    "Year"
    ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

```

&nbsp;

## 1) The Good

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# national belonging. 
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(national_belonging)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "National Belonging"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
  ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

```

## 2) The Bad

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# disinformation. 
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(disinformation)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Disinformation"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
    ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

# media distrust.  
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(media_distrust)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Media Distrust"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
  ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

# partisanship.   
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(partisanship)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Political Polarization"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
  ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

```

## 3) The Ugly

```{r, echo = FALSE, message = FALSE, warning = FALSE, dpi=600}

# discrimination.   
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(discrimination)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Social Discrimination"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
  ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

# racism.    
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(racism)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Racism"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
  ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

# sexism.     
sorted_data_frame %>% 
  mutate(
    z_score = get_z_scores(sorted_data_frame)
  ) %>%
  filter(
    term %in% str_to_title(sexism)
  ) %>%
  ggplot(
    aes(x=year, y=z_score, col=term)
  ) +
  geom_smooth(
    se=F, span = 0.5, size = 0.5
  ) +
  geom_vline(
    xintercept = 2016, col="darkgrey"
  ) +
  geom_vline(
    xintercept = 2020, col="darkgrey"
  ) +
  ggtitle(
    "Sexism"
  ) +
  ylab(
    "Z Score"
  ) +
  xlab(
    "Year"
  ) + 
  ylim(
    -2, 2.4
  ) +
  my_theme

```

&nbsp;

## **Sources**

- New York Times (2021) https://developer.nytimes.com/apis

- Statology (2021) https://www.statology.org/interpret-z-scores/

&nbsp;
&nbsp;
&nbsp;

